---
title: "Three exposure groups - long version"
lightbox: true
format:
  html:
    self-contained: true
    code-copy: true

execute:
  eval: true
  message: false
  warning: false
---


### Setting
Here is an example of a forest plot containing combined results of risk ratios for a comparison of three exposure groups (high dose, low dose, control, say) from a hypothetical study. These data are in long format, there is also a version in [wide format](ThreeGroupsV2.qmd). Some familiarity with the `forestploter` package is assumed. The reader might want to look at the [simple forest plot](OneSimplePlot.qmd) before continuing.

### Data

This version of the data is in long format, meaning each exposure group contributes with a row/observation for each subgroup that has been analyzed.


```{r}
library(forestploter)
library(grid)
library(data.table)
library(dplyr)

results <- fread("./data/TwoComparisonFPdataLong.csv")

# overview of selected variables - first 10 observations
print(results[1:10,c(1:4,6:8)], digits = 5)
```

The data needs a bit of editing before it is ready to be plugged into `forest`.

```{r}
round2 <- function(x) format(round(x,digits = 2), nsmall = 2, scientific = FALSE)

results <- results |>
  mutate(
    x = case_when(x == 0 ~ "Control",
                  x == 1 ~ "Low dose",
                  x == 2 ~ "High dose",
                  .default = ""),
    eventsn = case_when(is.na(events) ~ "",
                        .default = paste(events, "/", n)),
    riskratio = case_when(
      is.na(events) ~ "",
      is.na(rr) ~ "(ref)",
      .default = paste0(round2(rr), " (", round2(lcl), " - ", round2(ucl), ")"))
  ) |>
  select(level, x, eventsn, risk, riskratio, rr, lcl, ucl)

# empty spaces necessary for the forest function
results$blank <- "                          "
results$smallblank <- " "

colnames(results) <- c(
  "",
  "Exposure group",
  "Events / N",
  "Risk (95% CI)",
  "",
  "rr",
  "lcl",
  "ucl",
  "",
  ""
)

```

The data are ready to be plotted. In the theme we specify 4 colors for stripe-alternation; dark for subgroup header, white-light-white for specific exposure groups.

```{r}
tm <- forest_theme(
  base_size = 15,
  ci_pch = 19, 
  ci_lwd = 3,
  ci_col = "#58508d",
  refline_gp = gpar(col = "grey"),
  core = list(bg_params = list(fill = c("#ccc4e2",
                                        "white",
                                        "#f3eeff",
                                        "white"))) # 4 alternating colors of stripes
)

p <- forest(
  results[,c(1,2,10,3,10,4,10,5,9)],
  est = results$rr,
  lower = results$lcl,
  upper = results$ucl,
  ci_column = 9,
  ref_line = 1,
  xlim = c(.5, 1.5),
  ticks_at = c(.5, .75, 1, 1.25, 1.5),
  theme = tm
)

# add header for risk ratio / forest plot
p <- add_text(
  p,
  col = 8:9,
  row = 1,
  part = "header",
  text = "Risk ratio (95% CI)",
  gp = gpar(fontface = "bold", fontsize = 15)
)

# center aligning columns
p <- edit_plot(
  p,
  col = 4:8,
  part = "header",
  which = "text",
  hjust = unit(.5, "npc"),
  x = unit(.5, "npc")
)

p <- edit_plot(
  p,
  col = 4:8,
  part = "body",
  which = "text",
  hjust = unit(.5, "npc"),
  x = unit(.5, "npc")
)

# bold facing first column
p <- edit_plot(
  p,
  col = 1,
  part = "body",
  which = "text",
  gp = gpar(fontface = "bold")
)
```

```{r, echo = FALSE}
#| label: fig-fpthree1
#| fig-cap: |
#|   Forest plot with 3 exposure groups

p_wh <- get_wh(plot = p, unit = "in")
ggplot2::ggsave(
  "./figs/fpthree1.png",
  p,
  width = p_wh[1] * 1.05,
  height = p_wh[2],
  units = "in"
)

magick::image_read("./figs/fpthree1.png") |>
  magick::image_ggplot(interpolate = TRUE)
```

